---
title: "Generate bcerr comparisons"
author: "Laura White"
date: "2024-07-02"
output: html_document
---
```{r global options, include = FALSE}
knitr::opts_chunk$set(message=FALSE)
library(tidyverse)
library(cowplot)
library(ggtext)
library(ggrepel)
library(here)
here::i_am("Rmd/misc_generatebcerrfiles.Rmd")
```
This Rmarkdown document pulls in the set of tsvs containing basecalling error frequency generated using the python script `get_bcerror_freqs.py`, annotates those using sequence to structure information and modification data from the MODOMICS database, and writes everything out to a combined file, `combinedbcerr.tsv`.

Here's a function to read in the mappings from modified and unmodified tRNA aligned to a consensus sequence, for joining to sequence alignments.
```{r}
process_files <- function(prefix) {
  # Construct file paths
  mod_file_path <- here("ref/structural_alignments/modified", paste0(prefix, ".mods.seq2structure.tsv"))
  nomod_file_path <- here("ref/structural_alignments/unmodified", paste0(prefix, ".seq2structure.tsv"))

  # Read and process the modified file
  mod_map <- read_tsv(mod_file_path) %>%
    dplyr::rename(Position = seq_pos, mod_nt = struct_nt) %>%
    select(-struct_ref) %>%
    dplyr::rename(Reference = seq_ref) %>%
    distinct(Reference, Position, .keep_all = TRUE)

  # Read and process the unmodified file
  nomod_map <- read_tsv(nomod_file_path) %>%
    dplyr::rename(Position = seq_pos) %>%
    select(-struct_ref) %>%
    dplyr::rename(Reference = seq_ref) %>%
    distinct(Reference, Position, .keep_all = TRUE)

  # Join the datasets
  joined_map <- left_join(mod_map, nomod_map, by = c("Reference", "Position", "struct_pos"))

  # Add the 'species' column
  joined_map <- mutate(joined_map, species = prefix)

  # Assign the joined dataset to the global environment
  joined_map_name <- paste0(prefix, "_mappings")
  assign(joined_map_name, joined_map, envir = .GlobalEnv)
}
```

Below, bring in the mapping information.
```{r}
process_files("scerevisiae")
process_files("hsapiens")
process_files("tthermophila")
process_files("ecoli")
process_files("dmelanogaster")
bind_rows(scerevisiae_mappings, hsapiens_mappings, tthermophila_mappings, ecoli_mappings, dmelanogaster_mappings) -> mappings
rm(scerevisiae_mappings, hsapiens_mappings, tthermophila_mappings, ecoli_mappings, dmelanogaster_mappings)
```

Then bring in tsvs that contain the error frequencies for each sample.
```{r}
read_bcerr <- function(x, species, chemistry, bcmodel, bcversion) {
  df <- readr::read_tsv(x, show_col_types = FALSE)
  
  df <- df %>%
    dplyr::mutate(species = species,
                  chemistry = chemistry,
                  bcmodel = bcmodel,
                  bcversion = bcversion)
  
  return(df)
}

inputs <- list(
  list(file = here("bcerror/RNA004v5/sup/yeast/WTyeast004_20231111_1104_P2S-00519-B_PAQ47538_fa3726ec.rna004_130bps_sup@v5.0.0.tsv"), species = "scerevisiae", chemistry = "004", bcmodel = "sup", bcversion = "5"),
  list(file = here("bcerror/RNA004v5/sup/HumanFibroblast004_20231208_1351_MN42516_FAX71838_b8f7b990.rna004_130bps_sup@v5.0.0.tsv"), species = "hsapiens", chemistry = "004", bcmodel = "sup", bcversion = "5"),
  list(file = here("bcerror/RNA004v5/sup/Drosophila004_20231208_1307_MN31004_FAX73799_72c200e4.rna004_130bps_sup@v5.0.0.tsv"), species = "dmelanogaster", chemistry = "004", bcmodel = "sup", bcversion = "5"),
  list(file = here("bcerror/RNA004v5/sup/Ecoli004_20240105_1224_MN31004_FAX73799_1fe84761.rna004_130bps_sup@v5.0.0.tsv"), species = "ecoli", chemistry = "004", bcmodel = "sup", bcversion = "5"),
  list(file = here("bcerror/RNA004v5/sup/Tetrahymena004_20231208_0920_MN42516_FAX71838_a244513d.rna004_130bps_sup@v5.0.0.tsv"), species = "tthermophila", chemistry = "004", bcmodel = "sup", bcversion = "5"),
  
  list(file = here("bcerror/RNA004v5/hac/WTyeast004_20231111_1104_P2S-00519-B_PAQ47538_fa3726ec.rna004_130bps_hac@v5.0.0.tsv"), species = "scerevisiae", chemistry = "004", bcmodel = "hac", bcversion = "5"),
  list(file = here("bcerror/RNA004v5/hac/HumanFibroblast004_20231208_1351_MN42516_FAX71838_b8f7b990.rna004_130bps_hac@v5.0.0.tsv"), species = "hsapiens", chemistry = "004", bcmodel = "hac", bcversion = "5"),
  list(file = here("bcerror/RNA004v5/hac/Drosophila004_20231208_1307_MN31004_FAX73799_72c200e4.rna004_130bps_hac@v5.0.0.tsv"), species = "dmelanogaster", chemistry = "004", bcmodel = "hac", bcversion = "5"),
  list(file = here("bcerror/RNA004v5/hac/Ecoli004_20240105_1224_MN31004_FAX73799_1fe84761.rna004_130bps_hac@v5.0.0.tsv"), species = "ecoli", chemistry = "004", bcmodel = "hac", bcversion = "5"),
  list(file = here("bcerror/RNA004v5/hac/Tetrahymena004_20231208_0920_MN42516_FAX71838_a244513d.rna004_130bps_hac@v5.0.0.tsv"), species = "tthermophila", chemistry = "004", bcmodel = "hac", bcversion = "5"),
  
    list(file = here("bcerror/RNA004v3/WTyeast004_20231111_1104_P2S-00519-B_PAQ47538_fa3726ec.rna004_130bps_sup@v3.0.1.tsv"), species = "scerevisiae", chemistry = "004", bcmodel = "sup", bcversion = "3"),
  list(file = here("bcerror/RNA004v3/HumanFibroblast004_20231208_1351_MN42516_FAX71838_b8f7b990.rna004_130bps_sup@v3.0.1.tsv"), species = "hsapiens", chemistry = "004", bcmodel = "sup", bcversion = "3"),
  list(file = here("bcerror/RNA004v3/Drosophila004_20231208_1307_MN31004_FAX73799_72c200e4.rna004_130bps_sup@v3.0.1.tsv"), species = "dmelanogaster", chemistry = "004", bcmodel = "sup", bcversion = "3"),
  list(file = here("bcerror/RNA004v3/Ecoli004_20240105_1224_MN31004_FAX73799_1fe84761.rna004_130bps_sup@v3.0.1.tsv"), species = "ecoli", chemistry = "004", bcmodel = "sup", bcversion = "3"),
  list(file = here("bcerror/RNA004v3/Tetrahymena004_20231208_0920_MN42516_FAX71838_a244513d.rna004_130bps_sup@v3.0.1.tsv"), species = "tthermophila", chemistry = "004", bcmodel = "sup", bcversion = "3"),
  
  list(file = here("bcerror/RNA002/WTyeast002_20231111_1103_P2S-00519-A_PAS43478_1fa49e43.rna002_70bps_hac@v3.tsv"), species = "scerevisiae", chemistry = "002", bcmodel = "hac", bcversion = "1"),
  list(file = here("bcerror/RNA002/HumanFibroblast002_20231207_1455_P2S-00519-B_PAS29437_b7f6c4bd.rna002_70bps_hac@v3.tsv"), species = "hsapiens", chemistry = "002", bcmodel = "hac", bcversion = "1"),
  list(file = here("bcerror/RNA002/Drosophila002_20231207_1500_MN35252_FAX30455_48242de2.rna002_70bps_hac@v3.tsv"), species = "dmelanogaster", chemistry = "002", bcmodel = "hac", bcversion = "1"),
  list(file = here("bcerror/RNA002/Ecoli002_20231208_1312_P2S-00519-A_PAS44628_9e5214a7.rna002_70bps_hac@v3.tsv"), species = "ecoli", chemistry = "002", bcmodel = "hac", bcversion = "1"),
  list(file = here("bcerror/RNA002/Tetrahymena002_20231207_1743_MN35252_FAX30455_c9f80979.rna002_70bps_hac@v3.tsv"), species = "tthermophila", chemistry = "002", bcmodel = "hac", bcversion = "1")
)
# Apply the read_bcerr function to each input and combine the results
result_list <- lapply(inputs, function(input) {
  read_bcerr(input$file, input$species, input$chemistry, input$bcmodel, input$bcversion)
})

# Combine all data frames into one
allspecies <- dplyr::bind_rows(result_list)


# function to get 5mer sequence context for each mod
create_rolling_string <- function(nt_vec) {
  sapply(seq_along(nt_vec), function(i) {
    if (i < 3 | i > length(nt_vec) - 2) {
      return(NA)
    }
    
    rolling_elems <- c(lag(nt_vec, 2)[i], lag(nt_vec, 1)[i], nt_vec[i], lead(nt_vec, 1)[i], lead(nt_vec, 2)[i])
    
    if (any(is.na(rolling_elems))) {
      return(NA)
    } else {
      return(paste0(rolling_elems, collapse = ""))
    }
  })
}

left_join(allspecies, mappings, by = c("Reference", "Position", "species")) %>%
  select(-N_freq) %>%
  dplyr::rename(seq_pos = Position) %>%
  dplyr::rename(Position = struct_pos) %>%
  filter(!is.na(Position)) %>% # this removes tRNAs that do not have structural info in the mappings df
  mutate(Position = Position - 24) %>%
  group_by(species, Reference, chemistry, bcmodel, bcversion) %>%
  arrange(Position) %>%
  mutate(kmer_context = create_rolling_string(struct_nt)) %>%
  mutate(mod_context = create_rolling_string(mod_nt)) %>%
  filter(Bases_Mapped >29) -> mapped

# with ACGUTN intact
n_distinct(mapped$mod_context)
n_distinct(mapped$mod_nt)
```

We also pull in the database of Modomics codes in order to convert single character codes to short names.
Something here is wrong with the 5mer contexts.
```{r}
# Load and process the modomics data
modomics <- read_csv(here("Modomics/modomicscodes.csv")) %>%
  select(Name, `Short Name`, `RNAMods code (2023)`) %>%
  dplyr::rename(modname = Name, shortname = `Short Name`, mod_nt = `RNAMods code (2023)`) %>%
  group_by(mod_nt) %>%
  arrange(mod_nt, shortname)

# Remove duplicate mod_nt values from modomics
unique_modomics <- modomics %>%
  group_by(mod_nt) %>%
  slice(1) %>%  # Take the first entry for each mod_nt
  ungroup()

# Replace backslash with "Ħ" in the mapped dataset (old Modomics code usage)
mapped$mod_nt <- ifelse(mapped$mod_nt == "\\", "Ħ", mapped$mod_nt)

# Check distinct values before the join
distinct_mod_nt_before_join <- n_distinct(mapped$mod_nt)
print(paste("Distinct mod_nt before join:", distinct_mod_nt_before_join))

# Perform the left join with unique_modomics
mapped <- left_join(mapped, unique_modomics, by = "mod_nt")

# Check distinct values after the join
distinct_mod_nt_after_join <- n_distinct(mapped$mod_nt)
distinct_shortname_after_join <- n_distinct(mapped$shortname)
print(paste("Distinct mod_nt after join:", distinct_mod_nt_after_join))
print(paste("Distinct shortname after join:", distinct_shortname_after_join))

# save out mapped to bcerr
write_tsv(mapped, here("bcerror/combinedbcerr.tsv"))

# only keep modified positiosn with 30+ reads from the one dataset
# Ns are periods
mapped %>%
  filter(bcmodel == "sup" & bcversion == "5") %>%
  filter(Spanning_Reads > 29) %>%
  filter(!is.na(kmer_context) & !mod_nt %in% c('A', 'C', 'G', 'U', '.')) -> mapped

# after removing ACGUN do same counts of distinct mods, contexts
n_distinct(mapped$mod_context)
n_distinct(mapped$mod_nt)
```


