---
title: "Petit yeast candidate"
author: "Laura White"
date: "2024-01-10"
output: html_document
---
```{r global options, include = FALSE}
knitr::opts_chunk$set(message=FALSE)
library(tidyverse)
library(colorblindr)
library(cowplot)
library(stringr)
library(here)
here::i_am("Rmd/petit.Rmd")
```
We want to know if our mitochondrial tRNA aligning reads are genuine. So we've seqeunced tRNA from a petit yeast from EtBr treatment vs. its parent grande strain, and aligned these two to a reference containing both the mitochondrial and genomic yeast tRNAs.
```{r}
read_tRNA <- function(x, sample, ref) {
  df <- readr::read_tsv(x, col_names = c("chrom", "start", "pos", "score"), show_col_types = FALSE)
  
  df <- df %>%
    dplyr::mutate(sample = sample, ref = ref)
  
  return(df)
}

# both here is in case we want to compare to nuclear-only ref later
read_tRNA(here("bedgraphs/mito_and_nuc_refs/Grandeyeast004_20240105_1222_P2S-00519-B_PAS98845_231d7d7e.rna004_130bps_sup@v5.0.0.bwa_full_length.bam.fl.bg"), "grande", "both") -> grande
read_tRNA(here("bedgraphs/mito_and_nuc_refs/Petityeast004_20240105_1222_P2S-00519-A_PAQ47538_49891fce.rna004_130bps_sup@v5.0.0.bwa_full_length.bam.fl.bg"), "petit", "both") -> petit

combined <- list(grande, petit)


bind_rows(combined) %>%
  mutate(pos = as.numeric(pos)) %>%
  mutate(score = as.numeric(score)) %>%
  separate(chrom, c("origin", "tRNA", "AA", "anticodon", "family", "isotype")) %>%
  select(-tRNA) %>%
  mutate(anticodon = str_replace_all(anticodon, "U", "T")) -> tRNAcov # if any u and t inconsistency across refs


rm(grande, petit, combined)
```

So now we have normalized coverage on nuclear and mitochondrial tRNAs from the petit & grande strains.
```{r}
tRNAcov %>%
  group_by(sample, ref, origin, AA, anticodon, family, isotype) %>%
  slice_max(score, with_ties = F) %>%
  ungroup() %>%
  unite(tRNA, origin, AA, anticodon, sep = "-") %>%
  select(-pos) -> rollup
```

The code below adds up max scores for family & gene number info so that we have summary scores for each isodecoder.
```{r}
rollup %>%
  separate(tRNA, c("origin", "AA", "anticodon"), extra = "drop") %>%
  group_by(origin, AA, anticodon, sample) %>%
  summarize(score = sum(score)) %>%
  unite(tRNA, origin, AA, anticodon, sep = "-") %>%
  ungroup() %>%
  group_by(sample) %>%
  mutate(total_score = sum(score)) %>%
  mutate(CPM = score/total_score*1000000) -> isodecoder_scores
```

So one thing we want to ask is do the abundance scores for nuclear tRNAs change when we add the mitochondrial tRNAs to the reference.
Quick plot of everything aligned to the two references.
```{r}
isodecoder_scores %>%
  separate(tRNA, c("origin", "tRNA", extra="merge")) %>%
  filter(origin=="mito") %>%
  ggplot(., aes(y = CPM, x = tRNA, fill = fct_rev(sample))) +
  geom_bar(stat = "identity", width=.5, position = "dodge") +
  scale_fill_OkabeIto(order = c(2, 8)) +
  theme_cowplot() +
  theme(legend.position = 'right',
    legend.title = element_blank(),
    legend.direction ='vertical') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_text(margin = margin(r = 20))) +
  theme(aspect.ratio=1/3,
        plot.margin = margin(1, 1, 1, 1, "cm")) +
  labs(title = "Mitochondrial tRNA alignment rates",
       x = "",
       y = "Counts / Million") +
  coord_cartesian(expand = FALSE)

isodecoder_scores %>%
  separate(tRNA, c("origin", "tRNA", "anticodon")) %>%
  filter(origin=="nuc") %>%
  ggplot(., aes(y = CPM, x = tRNA, fill = fct_rev(sample))) +
  geom_bar(stat = "identity", width=.5, position = "dodge") +
  scale_fill_OkabeIto(order = c(2, 8)) +
  theme_cowplot() +
  theme(legend.position = 'right',
    legend.title = element_blank(),
    legend.direction ='vertical') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(aspect.ratio=1/3,
        plot.margin = margin(1, 1, 1, 1, "cm")) +
  labs(title = "Nuclear tRNA alignment rates",
       x = "",
       y = "Counts / Million") +
  coord_cartesian(expand = FALSE)
```
Roll up some statistics on how the abundances change
```{r}
isodecoder_scores %>%
  separate(tRNA, c("origin", "tRNA", extra="merge")) %>%
  group_by(origin, sample) %>%
  summarize(mean_CPM = mean(CPM, na.rm = T), .groups = 'drop') -> summary_data

# Calculate the fold change between grande and petit samples for each origin
fold_changes <- summary_data %>%
  spread(sample, mean_CPM) %>%
  mutate(fold_change = petit / grande)

# Print the summarized data and fold changes
print(summary_data)
print(fold_changes)

# Plot the differences
ggplot(differences, aes(x = origin, y = difference, fill = origin)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  theme_cowplot() +
  labs(title = "Difference in Mean CPM between Grande and Petit Samples",
       x = "Origin",
       y = "Difference in Mean CPM (Grande - Petit)") +
  scale_fill_OkabeIto(order = c(1, 2))
```


