---
title: "RNA_mods"
author: "Laura White"
date: "2024-01-14"
output: html_document
---
```{r global options, include = FALSE}
knitr::opts_chunk$set(message=FALSE)
library(tidyverse)
library(colorblindr)
library(ggokabeito)
library(cowplot)
library(ggbeeswarm)
library(gridExtra)
library(here)
here::i_am("Rmd/RNA_mods.Rmd")
```
We can take our tRNA basecalling error data and ask questions about how well we can detect different RNA modifications in multiple sequence contexts across multiple organisms. This is enabled by (a) abundant LC/MS data about tRNA modifications from Modomics and (b) our ability to pin our basecalling error data onto structure files and line them up with modification info.

Below, bring in the mapping information for budding yeast.
```{r}
read_tsv(here("ref/structural_alignments/modified/sacCer.mods.seq2structure.tsv")) %>%
  rename(Position = seq_pos, mod_nt = struct_nt) %>%
  select(-struct_ref) %>%
  rename(Reference = seq_ref) -> mod_mappings

read_tsv(here("ref/structural_alignments/unmodified/sacCer.seq2structure.tsv")) %>%
  rename(Position = seq_pos) %>%
  select(-struct_ref) %>%
  rename(Reference = seq_ref) -> nomod_mappings

left_join(nomod_mappings, mod_mappings) -> mappings
remove(mod_mappings, nomod_mappings)
```
Then bring in tsv that contains the error frequencies for budding yeast.
```{r}
read_bcerr <- function(x, Sample, ref) {
  df <- readr::read_tsv(x, show_col_types = FALSE)
  
  df <- df %>%
    dplyr::mutate(Sample = Sample)
  
  return(df)
}

# Function to create rolling strings of kmers
create_rolling_string <- function(nt_vec) {
  sapply(seq_along(nt_vec), function(i) {
    if (i < 3 | i > length(nt_vec) - 2) {
      return(NA)
    }
    
    rolling_elems <- c(lag(nt_vec, 2)[i], lag(nt_vec, 1)[i], nt_vec[i], lead(nt_vec, 1)[i], lead(nt_vec, 2)[i])
    
    if (any(is.na(rolling_elems))) {
      return(NA)
    } else {
      return(paste0(rolling_elems, collapse = ""))
    }
  })
}

read_bcerr(here("bcerror/WTyeast004_20231111.tsv"), "WT") -> wt

left_join(wt, mappings, by = c("Reference", "Position")) %>%
  select(-N_freq, -Position) %>%
  rename(Position = struct_pos) %>%
  filter(!is.na(Position)) %>% # this removes tRNAs that do not have structural info in the mappings df
  mutate(Position = Position -24) %>%
  group_by(Reference) %>%
  arrange(Position) %>%
  mutate(kmer_context = create_rolling_string(struct_nt)) %>%
  mutate(mod_context = create_rolling_string(mod_nt)) %>%
  filter(!is.na(kmer_context) & !mod_nt %in% c('A', 'C', 'G', 'T', 'U')) %>% # only keep positions with modifications
  filter(Coverage >29) %>% # enforce coverage threshold
  mutate(BCerror = MismatchFreq + InsertionFreq + DeletionFreq) -> mapped 

# how many unique mod sequence contexts? 267 for 27 modifications over almost 3k tRNA sites.
n_distinct(mapped$mod_context)
n_distinct(mapped$mod_nt)
```

We also pull in the database of Modomics codes in order to convert single character codes to short names.
```{r}
read_csv(here("Modomics/modomicscodes.csv")) %>%
  select(Name, `Short Name`, `RNAMods code (2023)`) %>%
  rename(modname = Name, shortname = `Short Name`, mod_nt = `RNAMods code (2023)`) -> modomics

left_join(mapped, modomics) -> mapped
```

```{r}
ggplot(mapped, aes(x = shortname, y = MismatchFreq)) +
  geom_jitter(cex = 0.2, alpha = 0.4, size = 0.5, width = 0.25) +  # Jitter plot
  geom_boxplot(width = 0.2, alpha = 0.5, outlier.shape = NA, fill = "#56B4E9") +
  theme_bw() +
  labs(title = "Mismatch frequency at modified nucleotides in budding yeast tRNAs",
       x = "",
       y = "Mismatch frequency") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.3))

```

Do that, but as a plotting function.
```{r}
plot_modification_data <- function(mapped, shortname) {
  # Filter for the specified modification using an explicit environment
  filtered_data <- mapped %>%
    filter(shortname == !!shortname)
  
 # Retrieve the corresponding mod_nt (they are 1 to 1, so just grab the first instance)
  mod_nt_value <- filtered_data$mod_nt[1]

  # Mismatch Frequency Plot
  plot_mismatch <- ggplot(filtered_data, aes(x = mod_context, y = MismatchFreq)) +
    geom_jitter(cex = 0.5, alpha = 1, size = 0.5, width = 0.25) +
    geom_boxplot(width = 0.2, alpha = .5, outlier.shape = NA, color = "#56B4E9") +
    theme_bw() +
    labs(title = paste("Mismatch frequencies in different nucleotide contexts,", shortname, "(",mod_nt_value,")"),
         x = "5mer nucleotide context",
         x = "5mer nucleotide context",
         y = "Mismatch frequency") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

  # Normalized Nucleotide Frequencies Plot
  normalized_data <- filtered_data %>%
    pivot_longer(cols = c(A_Freq, T_Freq, G_Freq, C_Freq), 
                 names_to = "Nucleotide", 
                 values_to = "Frequency") %>%
    group_by(mod_context) %>%
    mutate(Total_Freq = sum(Frequency, na.rm = TRUE)) %>%
    ungroup() %>%
    mutate(Normalized_Freq = Frequency / Total_Freq)

  plot_normalized <- ggplot(normalized_data, aes(x = mod_context, y = Normalized_Freq, fill = Nucleotide)) +
    geom_bar(stat = "identity", position = "fill") +
    theme_bw() +
    labs(title = paste("Normalized mismatch frequencies,", shortname, "modification", "(",mod_nt_value,")"),
         x = "5mer nucleotide context",
         x = "5mer nucleotide context",
         y = "Normalized nucleotide frequency") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    scale_fill_manual(values = c("A_Freq" = "#D55E00", "T_Freq" = "#0072B2", "G_Freq" = "#009E73", "C_Freq" = "#F0E442"))

  # Combine the plots side by side
  grid.arrange(plot_mismatch, plot_normalized, ncol = 2)
}

# Example usage of the function
plot_modification_data(mapped, "m1A")


```



To look more in depth at the flavor of basecalling errors, tidy dataset before plotting

```{r}
# Converting data to long format
long_data <- mapped %>%
  pivot_longer(cols = c(MismatchFreq, InsertionFreq, DeletionFreq), 
               names_to = "attribute", 
               values_to = "attribute_value")

# Creating faceted violin plot
ggplot(long_data, aes(x = attribute_value, y = attribute)) +
  geom_violin(trim = FALSE) +
  facet_wrap(~shortname, nrow =4) +
  theme_bw() +
  labs(title = "Basecalling error frequencies at yeast tRNA modifications",
       x = "",
       y = "Frequency")

```

