---
title: "Mod calls from BAMs"
author: "Laura White"
date: "2024-06-25"
output: html_document
---

```{r global options, include = FALSE}
knitr::opts_chunk$set(message=FALSE)
library(tidyverse)
library(Rsamtools)
library(cowplot)
library(here)
here::i_am("Rmd/modbams.Rmd")
```
I have a bunch of BAM files from SUP v5 calling with pseudouridine and m6A modification calls emitted into MM & ML tags. I want to look at these, both at the general distribution of modification likelihood (ML) scores, and then also the behavior at positions on tRNA that are annotated vs. unannotated for pseudouridylation in MODOMICS.

```{r}
# Function to extract ML values without extensive debugging information
extract_ml_values_single <- function(bam_file) {
  bam <- scanBam(BamFile(bam_file), param=ScanBamParam(tag=c("MM", "ML")))

  ml_values <- c()

  if (length(bam[[1]]$tag$MM) > 0) {
    for (i in seq_along(bam[[1]]$tag$MM)) {
      mm_tag <- bam[[1]]$tag$MM[[i]]
      ml_tag <- bam[[1]]$tag$ML[[i]]

      if (!is.null(mm_tag) && !is.null(ml_tag) && length(ml_tag) > 0) {
        mm_parts <- unlist(strsplit(mm_tag, ";"))
        ml_parts <- unlist(ml_tag)

        for (j in seq_along(mm_parts)) {
          if (grepl("T\\+17802", mm_parts[j])) {
            ml_values <- c(ml_values, as.numeric(ml_parts[j]))
          }
        }
      }
    }
  }

  return(ml_values)
}

# Path BAM file
single_bam_file <- "/Users/laurakwhite/Library/CloudStorage/Dropbox/Laura - UC Denver/2023/tRNAworkshop/alignments/004supv5/Tetrahymena004_20231208_0920_MN42516_FAX71838_a244513d.rna004_130bps_sup@v5.0.0.bwa_full_length.bam"

# Extract ML values from the single BAM file
ml_values <- extract_ml_values_single(single_bam_file)

if (length(ml_values) > 0) {
  ml_df <- data.frame(ML_values = ml_values, File = basename(single_bam_file))
  
  ggplot(ml_df, aes(x = ML_values)) +
    geom_histogram(binwidth = 5, fill = "blue", color = "black") +
    theme_minimal() +
    labs(title = "Distribution of ML Values for Pseudouridine, Tetrahymena", x = "ML Value", y = "Frequency")
} else {
  cat("No ML values found for file:", single_bam_file, "\n")
}
```
```{r}

# Step 1: Load the data
data <- read_tsv("/Users/laurakwhite/Library/CloudStorage/Dropbox/Laura - UC Denver/2023/tRNAworkshop/alignments/004supv5/subsampled.extract.tsv")

# Step 2: Filter for pseudouridine modifications
pseudoU_data <- data %>% filter(mod_code == "17802")

# Step 3: Calculate median modification quality scores
median_mod_qual <- pseudoU_data %>%
  group_by(ref_position) %>%
  summarize(median_mod_qual = median(mod_qual, na.rm = TRUE))

# Step 4: Plot the results
ggplot(median_mod_qual, aes(x = ref_position, y = median_mod_qual)) +
  geom_vline(xintercept = 79, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 54, linetype = "dashed", color = "red") +
  annotate("rect", xmin = -1, xmax = 23, ymin = -Inf, ymax = Inf, alpha = 0.2, fill = "gray") +
  geom_line() +
  geom_point() +
  labs(title = "Modification Quality Score for Pseudouridine (Ψ) Modifications",
       x = "Reference Position (all tRNAs)",
       y = "Median Modification Quality Score") +
  theme_cowplot()

```
```{r}
read_tsv(here("alignments/004supv5/yeastextract.tsv")) %>%
  filter(mod_code=="17802" & ref_strand=="+" & ref_mod_strand=="+" & mod_strand=="+" & mod_qual >.98) %>%
  select(read_id, ref_position, chrom, mod_qual, base_qual, ref_kmer) %>%
  mutate(ref_position = ref_position +1) -> test

read_tsv(here("alignments/004supv5/remorainfer.extract.tsv")) %>%
  filter(mod_code=="17802" & ref_strand=="+" & ref_mod_strand=="+" & mod_strand=="+") %>%
  select(read_id, ref_position, chrom, mod_qual, base_qual, ref_kmer) %>%
  mutate(ref_position = ref_position +1) -> remoratest

read_tsv(here("alignments/004supv5/mnremoved.remorainfer.extract.tsv")) %>%
  filter(mod_code=="17802" & ref_strand=="+" & ref_mod_strand=="+" & mod_strand=="+" & mod_qual >.98) %>%
  select(read_id, ref_position, chrom, mod_qual, base_qual, ref_kmer) %>%
  mutate(ref_position = ref_position +1) -> remoratest2
```
Let's start by getting data at one reference.
```{r}

data_filtered <- remoratest2 %>%
  filter(str_starts(chrom, "tRNA-Ala-AGC-1")) %>%
  filter(ref_position > 24 & ref_position < 101)

data_adjusted <- data_filtered %>%
  mutate(new_ref_position = ref_position - 24)

#create a complete set of adjusted reference positions
complete_positions <- data.frame(new_ref_position = seq(min(data_adjusted$new_ref_position), max(data_adjusted$new_ref_position)))

#merge the complete positions with the adjusted data
data_complete <- complete_positions %>%
  left_join(data_adjusted, by = "new_ref_position")

ggplot(data_complete, aes(x = as.factor(new_ref_position), y = mod_qual)) +
  geom_vline(xintercept = 50, linetype = "dashed", color = "red") +
  geom_jitter(width = 0.2, alpha = 0.2, color = "blue", na.rm = TRUE) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA, na.rm = TRUE) +
  theme_minimal() +
  labs(title = "Modification probabilities at all pseudoU sites on Ala-AGC-1",
       x = "Position",
       y = "Modification Quality") +
  scale_x_discrete(drop = FALSE) +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust =.5))

```
Generated pileup data at 0.98 threshold, loading this in with column names
```{r}
# Define the column names
col_names <- c("chrom", "start", "end", "fmod", "ncov")

# Read the bedgraph file with specified column names
yeastpileup <- read_tsv(here("alignments/004supv5/bedgraphs/yeast/17802_positive.bedgraph"), col_names = col_names)

yeastpileup %>%
  select(-end) %>%
  mutate(start = start-24) %>%
  dplyr::rename(position = start) %>%
  filter(str_starts(chrom, "tRNA-Ala-AGC-1")) -> alatRNA

#All the alanine tRNAs above have the same positional info, so we can collpase this data
collapsed_data <- alatRNA %>%
  group_by(position) %>%
  summarise(
    weighted_fmod = sum(fmod * ncov),
    total_ncov = sum(ncov)
  ) %>%
  mutate(final_fmod = weighted_fmod / total_ncov) %>%
  select(position, final_fmod, total_ncov) %>%
  dplyr::rename(fmod = final_fmod, ncov = total_ncov)

ggplot(collapsed_data, aes(x = position, y = fmod)) +
  geom_vline(xintercept = 79 - 24, linetype = "dashed", color = "red") +
  geom_jitter(width = 0.2, alpha = 0.2, color = "blue", na.rm = TRUE) +
#  geom_boxplot(alpha = 0.6, outlier.shape = NA, na.rm = TRUE) +
  theme_minimal() +
  labs(title = "Modification probabilities at high confidence pseudoU sites on Ala-AGC-1",
       x = "Position",
       y = "Modification Quality") +
  scale_x_discrete(drop = FALSE) +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


# Create a complete sequence of positions from min to max
complete_positions <- tibble(position = seq(min(collapsed_data$position), max(collapsed_data$position)))

# Join the complete sequence with your dataset
complete_data <- complete_positions %>%
  left_join(collapsed_data, by = "position")

# Create two separate data frames
all_data <- complete_data %>%
  mutate(facet = "All Data")

filtered_data <- complete_data %>%
  filter(ncov > 29) %>%
  mutate(facet = "ncov > 29")

# Combine the two data frames
combined_data <- bind_rows(all_data, filtered_data)

# Plot the data
ggplot(combined_data, aes(x = position, y = fmod)) +
  geom_col() +
  scale_x_continuous(breaks = seq(min(combined_data$position), max(combined_data$position), by = 1)) +
  labs(title = "Fraction modified on Ala-AGC, all mod calls",
       x = "Position",
       y = "Fraction modified") +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  facet_wrap(~ facet, scales = "free_y", ncol = 1) +
  geom_hline(yintercept = .5, linetype = "dashed", color = "red")
```
A retest using the bedmethyl output
```{r}
read_tsv(here("alignments/004supv5/bedmethyls/WTyeastremorainfer.98.out.bed")) %>%
  filter(name == "17802" & strand == "+") %>%
  select(chrom, chromStart, valid_coverage, percent_modified, count_modified,  count_other_mode, count_canonical, count_delete, count_fail, count_diff, count_nocall) %>%
  dplyr::rename(start = chromStart) %>%
  mutate(count_unmodified = count_canonical + count_other_mode + count_fail + count_nocall + count_diff + count_delete) %>%
  mutate(count_all = count_modified + count_unmodified) %>%
  mutate(scaled_mod_pct = (count_modified / count_all)*100) -> yeastbedmeth

yeastbedmeth %>%
  ggplot(aes(x = percent_modified, y = scaled_mod_pct)) +
  geom_point(alpha = .2) +
  theme_cowplot() +
  labs(title = "What does percent modified mean at --filter-threshold 0.9? \n each dot is one position mapped to the reference",
       x = "ModKit % modified in pileup bedmethyl column",
       y = "Mod calls as % of all reads spanning position")


yeastbedmeth %>%
  filter(count_all > 29) %>%
  ggplot(aes(x = percent_modified, y = scaled_mod_pct)) +
  geom_point(alpha = .2) +
  theme_cowplot() +
  labs(title = "What does percent modified mean at --filter-threshold 0.9? \n each dot is one position mapped to the reference \n with total read coverage >29",
       x = "ModKit % modified in pileup bedmethyl column",
       y = "Mod calls as % of all reads spanning position")
```
```{r}
yeastbedmeth %>%
  mutate(start = start-23) %>%
  dplyr::rename(position = start) %>%
  filter(str_starts(chrom, "tRNA-Ala-AGC-1")) -> alatRNAbedmeth

# Step 1: Summarise the data to calculate sums
summarised_data <- alatRNAbedmeth %>%
  group_by(position) %>%
  summarise(
    sum_percent_modified = sum(percent_modified * count_all, na.rm = TRUE),
    total_coverage = sum(count_all, na.rm = TRUE),
    modkitcov = sum(count_modified + count_canonical)
  )

# Step 2: Calculate the weighted percent modified
collapsed_data <- summarised_data %>%
  mutate(
    fmod = sum_percent_modified / total_coverage,
    ncov = total_coverage
  ) %>%
  select(position, fmod, ncov, modkitcov)

# Create a complete sequence of positions from min to max
complete_positions <- tibble(position = seq(min(collapsed_data$position), max(collapsed_data$position)))

# Join the complete sequence with the collapsed data
complete_data <- complete_positions %>%
  left_join(collapsed_data, by = "position")

# Filter the data for ncov > 29
filtered_data <- complete_data %>%
  filter(modkitcov > 29)

# Plot the data
ggplot(filtered_data, aes(x = position, y = fmod)) +
  geom_col() +
  scale_x_continuous(breaks = seq(min(filtered_data$position), max(filtered_data$position), by = 1)) +
  labs(title = "ModKit percent modified on Ala-AGC-1",
       x = "Position",
       y = "Percent modified") +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))



# Step 2: Calculate the weighted percent modified
alatRNAbedmeth %>%
  group_by(position) %>%
  summarise(
    scaled_mod_pct = sum(scaled_mod_pct * count_all, na.rm = TRUE),
    total_coverage = sum(count_all, na.rm = TRUE)
  ) %>%
  mutate(
    fmod = scaled_mod_pct / total_coverage,
    ncov = total_coverage
  ) %>%
  select(position, fmod, ncov) %>%
  filter(ncov > 29) -> scaleddata
  

ggplot(scaleddata, aes(x = position, y = fmod)) +
  geom_col() +
  scale_x_continuous(breaks = seq(min(scaleddata$position), max(scaleddata$position), by = 1)) +
  labs(title = "Coverage-scaled percent modified on Ala-AGC-1",
       x = "Position",
       y = "Percent modified") +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))
```

